---

title: Race & Sex Estimates in R
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_notebook
---

This gist demonstrates how to estimate the race and sex of identified individuals using their name and location. 

## Loading packages

When installing packages, I've really been enjoying the `pacman` package lately. The `p_load` function will load your packages and install them if they are not already. 

```{r}

if(!require(pacman)) install.packages("pacman")
pacman::p_load(
			   randomNames, # random name generator for the last example
			   readr, # for reading my hidden api keys stored on my machine
			   ggmap, # geocoding using Google's API
			   gender, # estimate sex using first names
			   wru, # estimate rate using location and surname
			   tigris, # pull just census shapefiles
			   sf, # spatial simple features - a better use than shapefiles
			   leaflet, # other mapping features
			   tmap, # interactive mapping
			   tidycensus, # pulling various census data and shapefiles
			   tidyverse # various packages using tidyverse
			)
```

The old-school method, which doesn't install the packages for you, is: 

```{r}
library(randomNames) 
library(readr)  
library(ggmap) 
library(gender) 
library(wru) 
library(tigris) 
library(sf) 
library(tidyverse)
```

To avoid getting a bunch of API keys for Google maps and the census, just load this file

```{r} 
load("~/git/git_tutorials/race_sex_estimates/tutorial_data.RData")
```
## Example 1: Simple estimations

### Create a simple dataset

To get a brief look under the hood of the race and sex estimates, we're going to start with a simple dataset that I created for the Oakland, CA area. These are names I made up with random, real addresses I selected using google maps. 

```{r}
	firstname <- c("John","Nia","Lupe","Conner","Jamal")
	surname <- c("Smith","Carter","Rodriguez","Miller","Williams")
	age <- c(30, 30, 30, 30, 30) # if you don't have age, use 30
	address <- c("407 Perkins St, Oakland, CA 94610","1531 E 15th St, Oakland, CA 94606","752 Alcatraz Ave, Oakland, CA 94609","1198 12th St, Oakland, CA 94607","630 Foothill Blvd, Oakland, CA 94606")
	df <- data.frame(firstname,surname,age, address, stringsAsFactors = FALSE)
	df$id <- rownames(df)
```

This is how your dataframe should look when you're preparing it for estimation. 

```{r}
	df
```
### Geocode our data

```{r}
	# Beginning June 11, 2018, Google requires an API key. Go here to register for an API key (requires credit card but your trial should cover most geocodes)
	# https://developers.google.com/maps/documentation/javascript/get-api-key#step-1-get-an-api-key-from-the-google-cloud-platform-console

	# Register your API
	register_google(key = read_file("~/.ssh/googleapikey.txt")) # use your own google api here
	getOption("ggmap")

	# Geocode
	geo <- geocode(df$address, output = "more")

	geo

		# Pay attention to the loctype, "approximation" is very
		# braod and not a good geocode.
		# read: http://www.nickeubank.com/wp-content/uploads/2015/10/RGIS4_geocoding.html#loctype

	alameda_tracts <- 
		tracts(state = "CA", # get tract fips for each point
			   county = "Alameda", 
			   year = 2017, 
			   class = "sf", 
			   cb = TRUE) 

	# Combine names and locations
	df_geo <- cbind(df,geo) %>% 
			  st_as_sf(coords = c("lon", "lat")) %>%
			  st_set_crs(4326) %>%
			  st_transform(4269) %>%
			  st_join(., alameda_tracts)

```

### Estimate sex

```{r}
	# Estimate names
	sex <- gender(unique(df_geo$firstname),
				  method = "ssa",
				  countries = "United States") %>%
		   data.frame() %>%
		   distinct()

	# Join sex probabilities to
	df_geo_sex <- left_join(df_geo,
							sex,
							by = c("firstname" = "name"))

	# see which names were unidentifiable
	df_geo_sex %>%
	filter(is.na(gender)) %>%
	data.frame()
		# Up to you to decide what you do with these. One method I used was to
		# search facebook for these names and take the first 10 profiles I found
		# and averaged the gender based on these 10
```

### Estimate race

```{r}
		# You'll need a US Census API. Go here to sign up for one
		# https://api.census.gov/data/key_signup.html

	tract.data <-
		get_census_data(key = read_csv("~/.ssh/censusapi.csv")$key,
						"CA",
      					census.geo = "tract")

	race <- df_geo_sex %>%
			select(id,
				   surname,
				   county = COUNTYFP,
				   tract = TRACTCE) %>%
			mutate(state = "CA") %>%
			st_set_geometry(NULL) %>%
			predict_race(.,
				 		census.geo = "tract",
				 		census.data = tract.data) %>% 
			arrange(id) %>%
			rename(white = pred.whi, # for easier interpretation
				   black = pred.bla,
				   latinx = pred.his,
				   asian = pred.asi,
				   other = pred.oth) %>%
			mutate(race = colnames(.[,6:10])[max.col(.[,6:10],
										ties.method="first")])

	df_geo_sex_race <- left_join(df_geo_sex, race %>% select(id, surname, race))
```

### Map these individuals

```{r}
tmap_mode("view")

tm_basemap(leaflet::providers$CartoDB.Positron) + 
	# For other basemaps see: http://leaflet-extras.github.io/leaflet-providers/preview/
	tm_shape(df_geo_sex_race, name = "Estimated Race") +
	tm_symbols(shape = "race", 
			   col = "race",
			   popup.vars = c("First Name" = "firstname",
			   				  "Last Name" = "surname", 
			   				  "Sex" = "gender", 
			   				  "Race" = "race"))
```

## Example 2: Larger estimates

```{r}
r_names <- randomNames(10000)

TRACTCE <- sample(alameda_tracts$TRACTCE, 
					size = 10000, 
					replace = TRUE)	

big_df <- 
	data.frame(r_names, TRACTCE) %>%
	separate(r_names, c("surname", "firstname"), 
			 sep = ", ") %>% 
	mutate(STATEFP = "06", 
		   COUNTYFP = "001", 
		   id = rownames(.))

big_sex <- 
	gender(unique(big_df$firstname),
				  method = "ssa",
				  countries = "United States") %>%
	data.frame() %>%
	distinct()

big_sex_df <- 
	left_join(big_df,
				big_sex,
				by = c("firstname" = "name"))
	
	big_sex_df %>% 
	filter(is.na(gender)) %>%
	glimpse()

big_race <- big_sex_df %>% 
			select(id,
				   surname,
				   county = COUNTYFP,
				   tract = TRACTCE) %>% 
			mutate(state = "CA") %>% 
			predict_race(.,
				 		census.geo = "tract",
				 		census.data = tract.data) %>% 
			arrange(id) %>% 
			rename(white = pred.whi, # for easier interpretation
				   black = pred.bla,
				   latinx = pred.his,
				   asian = pred.asi,
				   other = pred.oth) %>% 
			mutate(race = colnames(.[,6:10])[max.col(.[,6:10],
										ties.method="first")])

big_sex_race <- left_join(big_sex_df, big_race %>% select(id, surname, white:race))

ind <- 
    big_sex_race %>% 
    group_by(STATEFP, COUNTYFP, TRACTCE) %>% 
    mutate(
        white_male = white*proportion_male, 
        white_female = white*proportion_female, 
        white_unknown = case_when(is.na(gender) ~ white),
        black_male = black*proportion_male, 
        black_female = black*proportion_female, 
        black_unknown = case_when(is.na(gender) ~ black),
        latinx_male = latinx*proportion_male, 
        latinx_female = latinx*proportion_female, 
        latinx_unknown = case_when(is.na(gender) ~ latinx),
        asian_male = asian*proportion_male, 
        asian_female = asian*proportion_female, 
        asian_unknown = case_when(is.na(gender) ~ asian),
        other_male = other*proportion_male, 
        other_female = other*proportion_female,
        other_unknown = case_when(is.na(gender) ~ other)
    ) %>% 
    summarise(
        tr_total = sum(white,black,latinx,asian,other, na.rm = TRUE), 
        tr_male = sum(proportion_male, na.rm = TRUE), 
        tr_female = sum(proportion_female, na.rm = TRUE), 
        tr_white = sum(white, na.rm = TRUE), 
        tr_black = sum(black, na.rm = TRUE), 
        tr_latinx = sum(latinx, na.rm = TRUE), 
        tr_asian = sum(asian, na.rm = TRUE), 
        tr_other = sum(other, na.rm = TRUE),
        tr_white_male = sum(white_male, na.rm = TRUE), 
        tr_white_female = sum(white_female, na.rm = TRUE), 
        tr_white_unknown = sum(white_unknown, na.rm = TRUE), 
        tr_black_male = sum(black_male, na.rm = TRUE), 
        tr_black_female = sum(black_female, na.rm = TRUE), 
        tr_black_unknown = sum(black_unknown, na.rm = TRUE), 
        tr_latinx_male = sum(latinx_male, na.rm = TRUE), 
        tr_latinx_female = sum(latinx_female, na.rm = TRUE), 
        tr_latinx_unknown = sum(latinx_unknown, na.rm = TRUE), 
        tr_asian_male = sum(asian_male, na.rm = TRUE), 
        tr_asian_female = sum(asian_female, na.rm = TRUE), 
        tr_asian_unknown = sum(asian_unknown, na.rm = TRUE), 
        tr_other_male = sum(other_male, na.rm = TRUE), 
        tr_other_female = sum(other_female, na.rm = TRUE),
        tr_other_unknown = sum(other_unknown, na.rm = TRUE))

 big_sex_race_sf <- 
 	left_join(alameda_tracts, ind)
```

### Map these polygons

```{r}
tm_basemap(leaflet::providers$CartoDB.Positron) + 
	# For other basemaps see: http://leaflet-extras.github.io/leaflet-providers/preview/
	tm_shape(big_sex_race_sf, name = "Estimated Race") +
	tm_polygons(col = c("tr_white", "tr_black","tr_latinx","tr_asian","tr_other"), 
				alpha = .5)
```


```{r}
save.image("~/git/git_tutorials/race_sex_estimates/tutorial_data.RData")
```
